Combining global tree cover loss data with national historical forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar

### 2017-04-21 deforestmap.Rmd
Comparison

Accurate maps of forest-cover and forest-cover change are necessary to monitore carbon emissions and loss of biodiversity associated to deforestation, assess the efficiency of present conservation strategies [@Eklund2016], and implement new strategies for the future [@Vieilledent2013a, @Vieilledent2016]. Unfortunately, accurate and cloud-free forest-cover maps are not available for Madagascar for the last decade (2005-2015). In their paper untitled _"Fifty years of deforestation and forest fragmentation in Madagascar"_, @Harper2007 published maps of forest cover change for Madagascar from 1950s to 2000 (for the years _c._ 1953, _c._ 1973, 1990 and 2000). The _c._ 1953 forest map was derived from the visual interpretation of aerial photographies and gives a coarse representation (scale 1/1,000,000) of the forest extent in Madagascar for the 1950s. Forest maps for the years _c._ 1973, 1990 and 2000, were obtained from a supervised classification of Landsat satellite images at 30m and can be used to derive much more accurate estimates of the forest cover at these three dates. Maps provided by @Harper2007 were not free of clouds (11 244 km2 of unknown cover type for the year 2000 for example). Using the same supervised classification approach as in @Harper2007, more recent maps have been released for the periods 2000-2005-2010 by national institutions, with the technical support of international environmental NGOs [@MEFT2009, @ONE2013]. Another recent map using more advanced statistical tool for classification (the Random Forest classifier, see @Grinand2013 and @Rakotomalala2015) has been released for the periods 2005-2010-2013 [@ONE2015]. However, these maps (i) are too old to give recent estimates of deforestation [@MEFT2009, @ONE2013], (ii) figure large areas of clouds (images with more than 20% of clouds in 2010 for the map provided by [@ONE2013]), or (iii) show large errors of classification in specific areas, especially in the dry and spiny forests for which the spectral answer varies strongly due to deciduousness associated with the higher seasonality than in moist forests (Kappa<0.6 and overall accuracy<0.8 for the dry and spiny forests for the maps produced by @ONE2015). Moreover, deriving such forest maps from supervised classification requires a huge amount of time and work, especially regarding image selection to reduce cloud cover and visual interpretation of the numerous observation polygons used to train the classification algorithm [@Rakotomalala2015]. In addition, visual interpretation might be subjective and is potentially a source of errors. All this work of image selection and visual interpretation have to be repeated to produce new forest maps in the future.

@@

We derived 30m cloud-free forest maps for Madagascar for the years 1990, 2000, 2010 and 2014 combining the 1990 and 2000 forest maps by @Harper2007 and the 2000-2014 tree cover and tree cover loss maps by @Hansen2013. On the 2000 Harper's map, 200,000 ha clouds are present over the 4.2 million ha moist forest ecoregion (4.8% of clouds). To remove these clouds, we used the 2000 cloud-free tree cover map provided by @Hansen2013 (also at 30m resolution) and we chose a threshold of 75% of tree cover [@Achard2014] to decide whether to replace clouds with forest or non-forest. We thus obtained a cloud-free year 2000 forest-cover map for Madagascar. From this map, using tree cover loss data from 2000 to 2014 provided by @Hansen2013, we obtained annual cloud-free forest-cover maps from 2010 to 2014 at 30m resolution. We corrected the cloudy 1990 Harper forest map assuming that if forest was present in 2000, it was also present in 1990. Doing so, cloud-cover was reduced to a negligeable quantity of `r epsilon_1990`ha. For these residual cloud areas, we assumed that forest-cover was the same as for the year 2000. We also corrected the cloudy 1973 Harper forest map assuming that if forest was present in 1973, it was also present in 1973. Contrary to the year 1990, we were not able to remove all the clouds on the 1973 map and the remaining cloud cover was of `r round(cloud_1973/10e6,2)` million ha. Because the 1953 forest-cover map was obtained from the digitalization of a paper map derived from aerial photographies, and thus cannot be perfectly aligned with the maps obtained in the following years, we did not performed any modifications on the original forest map of 1953 except reprojection. For all forest-cover maps from 1973, we removed the isolated non-forest pixels if all the surrounding pixels were covered by forest, assuming that small-scaled deforestation on a 30m square pixel inside the forest was unlikely. This was important in order not to over-estimate forest fragmentation. All these maps are made freely available at https://bioscenemada.cirad.fr/forestmaps.

### 2.2 Computing forest-cover area and deforestation rate and comparing estimates with previous studies

On the basis of the forest-cover maps we obtained, we computed the forest-cover area (in ha) and the mean annual deforestation rate (in hectares of forest and in percentage) on six decades, for five time periods from _c._ 1953 to 2014 (_c._ 1953-_c._ 1973-1990-2000-2010-2014). Mean annual deforestation rates in percentage were computed as follow [@Puyravaud2003]: $\theta = 100 \times [1-(1-(F_{t2}-F_{t1})/F_{t1})^{(1/(t2-t1))}]$, $\theta$ being the annual deforestation rate (in %/yr), $F_{t2}$ and $F_{t1}$ being the forest covers (reported in Tab. 2) at times $t2$ and $t1$ (in years). Because of the large cloud-cover in 1973, the annual deforestation rates in hectares of forest for the two periods _c._ 1953-1973 and 1973-1990 are only rough estimates of the deforestation rates. This problem was overcome when computing the annual deforestation rates in percentage for these two periods. In this case, we only considered the cover change and computed the ratio $F_{t2}-F_{t1})/F_{t1}$ for pixels that were not covered by clouds in 1973. Computation were done at the national scale and for each forest type in Madagascar: the moist forest in the East, the dry forest in the West, the spiny forest in the South and mangroves on the West coast. To define the forest type, we used a map provided by the MEFT (_"Ministère de l'Environnement et des Forêts à Madagascar"_) showing the boundaries of the four ecoregions in Madagascar. Ecoregions were defined on the basis of climatic and vegetation criteria using the climate classification by @Cornet1974 and the vegetation classification from the 1996 IEFN national forest inventory [@IEFN1996].

We compared our estimates of forest-cover and forest-cover change with three previous studies that have estimated forest-cover and forest-cover change at the national scale for Madagascar [@Harper2007; @MEFT2009; @ONE2015]. @Harper2007 provided estimates for the periods _c._ 1953-_c._ 1973-1990-2000, @MEFT2009 provided estimates for the periods 1990-2000-2005 and @ONE2015 provided estimates for the periods 2005-2010-2013. The same ecoregion boundaries as in our study have been used in the @ONE2015 report but this was not the case for the @Harper2007 and @MEFT2009 reports, which can explain some of the differences between studies in the forest-cover areas and deforestation rates by ecoregions. W

@@

```{r defor, results="asis"}
perc.1 <- round(100*(defor.df$area[defor.df$Year==1973]-defor.df$area[defor.df$Year==2014])/defor.df$area[defor.df$Year==1973])
perc.2 <- round(100*(defor.df$area[defor.df$Year==1953]-defor.df$area[defor.df$Year==2014])/defor.df$area[defor.df$Year==1953])
d.1 <- defor.df$theta[defor.df$Year==1990]
d.2 <- defor.df$theta[defor.df$Year==2010]
tab.1 <- defor.df
names(tab.1) <- c("Year", "Forest (ha)", "Clouds (ha)", "Annual defor. (ha/yr)", "theta (%/yr)")
kable(tab.1)
```

@@

```{r fcc_comp, results="asis"}
tab.2 <- df.comp
tab.2 <- tab.2[,c(2,1,3:10)]
tab.2$ForestType <- c("Total",rep("",3),"Moist",rep(" ",3),"Dry",rep(" ",3),"Spiny",rep(" ",3),"Mangroves",rep(" ",3))
col.names <- c("Forest type","Source","1953","1973","1990","2000","2005","2010","2013","2014")
kable(tab.2,col.names=col.names,align=c("l","l",rep("r",8)))
#print(xtable(tab.2,col.names=col.names,align=c("llrrrrrrrrr")),type="html")
```

@@

Table 3: **Comparison of Madagascar annual deforestation rates with previous studies on the period 1953-2014**. Annual deforestation rates are providing in ha/yr and in %/yr (second number in parenthesis). For our study, we used the following formula [@Puyravaud2003] to compute the annual deforestation rates: $\theta = 100 \times [1-(1-(F_{t2}-F_{t1})/F_{t1})^{(1/(t2-t1))}]$, with $\theta$ the annual deforestation rate (in %/yr), $F_{t2}$ and $F_{t1}$ the forest covers (reported in Tab. 2) at times t2 and t1 (in years).

```{r frag, results="asis"}
tab.4 <- frag.df
p.1 <- round(frag.df$interior[frag.df$Year==1973])
p.2 <- round(frag.df$interior[frag.df$Year==2014])
p.3 <- round(frag.df$patch[frag.df$Year==2014]+frag.df$transitional[frag.df$Year==2014])
names(tab.4) <- c("Year","Forest (ha)", "patch (%)", "transitional (%)", "edge (%)", "perforated (%)", "interior (%)", "NA (%)")
kable(tab.4)
```

>>>>

Accurate maps of forest-cover and forest-cover change are necessary to monitore carbon emissions and loss of biodiversity associated to deforestation, assess the efficiency of present conservation strategies [@Eklund2016], and implement new strategies for the future [@Vieilledent2013a, @Vieilledent2016]. Unfortunately, accurate and cloud-free forest-cover maps are not available for Madagascar for the last decade (2005-2015). In their paper untitled _"Fifty years of deforestation and forest fragmentation in Madagascar"_, @Harper2007 published maps of forest cover change for Madagascar from 1950s to 2000 (for the years _c._ 1953, _c._ 1973, 1990 and 2000). The _c._ 1953 forest map was derived from the visual interpretation of aerial photographies and gives a coarse representation (scale 1/1,000,000) of the forest extent in Madagascar for the 1950s. Forest maps for the years _c._ 1973, 1990 and 2000, were obtained from a supervised classification of Landsat satellite images at 60 (for the year 1973) and 30m (for years 1990 and 2000) and can be used to derive much more accurate estimates of the forest cover at these three dates. Maps provided by @Harper2007 were not free of clouds (11 244 km2 of unknown cover type for the year 2000 for example). Using the same supervised classification approach as in @Harper2007, more recent maps have been released for the periods 2000-2005-2010 by national institutions, with the technical support of international environmental NGOs [@MEFT2009, @ONE2013]. Another recent map using more advanced statistical tool for classification (the Random Forest classifier, see @Grinand2013 and @Rakotomalala2015) has been released for the periods 2005-2010-2013 [@ONE2015]. However, these maps (i) are too old to give recent estimates of deforestation [@MEFT2009, @ONE2013], (ii) figure large areas of clouds (images with more than 20% of clouds in 2010 for the map provided by [@ONE2013]), or (iii) show large errors of classification in specific areas, especially in the dry and spiny forests, for which the spectral answer varies strongly due to deciduousness associated with the higher seasonality than in moist forests (Kappa<0.6 and overall accuracy<0.8 for the dry and spiny forests for the maps produced by @ONE2015). Moreover, deriving such forest maps from supervised classification requires a huge amount of time and work, especially regarding image selection to reduce cloud cover and visual interpretation of the numerous observation polygons used to train the classification algorithm [@Rakotomalala2015]. In addition, visual interpretation might be subjective and is potentially a source of errors. All this work of image selection and visual interpretation have to be repeated to produce new forest maps in the future.

@@

We derived 30m cloud-free annual forest maps for Madagascar for the period 2000-2014 combining the 2000 forest map by @Harper2007 and the 2000-2014 tree cover and tree cover loss maps by @Hansen2013. On the 2000 Harper's forest map, `r ha.cloud.2000` ha of clouds are present, mostly (`r perc.cloud.moist`) over the `r round(ha.forest.2000.moist/1e6,2)` million ha moist forest ecoregion. To remove these clouds, we used the 2000 cloud-free tree cover map provided by @Hansen2013 (at 30m resolution) and we chose a threshold of 75% of tree cover [@Achard2014] to decide whether to replace clouds with forest or non-forest. We thus obtained a cloud-free year 2000 forest-cover map for Madagascar. From this map, using tree cover loss data from 2000 to 2014 provided by @Hansen2013, we obtained annual cloud-free forest-cover maps from 2010 to 2014 at 30m resolution. We then corrected the cloudy 1990 Harper forest map assuming that if forest was present in 2000, it was also present in 1990. Doing so, cloud-cover was reduced to a negligeable quantity of `r residual.clouds.1990` ha. For these residual cloud areas, we assumed that forest-cover was the same as for the year 2000. We also corrected the cloudy 1973 Harper forest map assuming that if forest was present in 1990, it was also present in 1973. Contrary to the year 1990, we were not able to remove all the clouds on the 1973 map and the remaining cloud cover was of `r round(cloud.1973/1e6,2)` million ha. Because the 1953 forest-cover map was obtained from the digitalization of a paper map derived from aerial photographies, and thus cannot be perfectly aligned with the maps obtained in the following years, we did not performed any modifications on the original forest map of 1953 except reprojection. For all forest-cover maps from 1973, we removed the isolated non-forest pixels if all the surrounding pixels were covered by forest, assuming that small-scaled deforestation on a 30m square pixel inside the forest was unlikely. This was important in order not to over-estimate forest fragmentation. All these maps are made freely available at https://bioscenemada.cirad.fr/forestmaps.

### 2.2 Computing forest-cover area and deforestation rate

On the basis of the forest-cover maps we obtained, we computed the forest-cover area (in ha) and the mean annual deforestation rate (in ha/yr and in %/yr) on six decades, for five time periods from _c._ 1953 to 2014 (_c._ 1953-_c._ 1973-1990-2000-2010-2014). Mean annual deforestation rates in percentage were computed as follow [@Puyravaud2003; @Vieilledent2013]: $\theta = 100 \times [1-(1-(F_{t2}-F_{t1})/F_{t1})^{(1/(t2-t1))}]$. In this formula, $\theta$ is the annual deforestation rate (in %/yr), $F_{t2}$ and $F_{t1}$ are the forest covers free of clouds at both dates $t2$ and $t1$, and $t2-t1$ is the time-interval (in years) between the two dates. Because of the large cloud-cover (`r round(cloud.1973/1e6,2)` million ha) in 1973, the annual deforestation rates in hectares of forest for the two periods _c._ 1953-1973 and 1973-1990 are only rough estimates of the deforestation rates in number of hectares. This problem was overcome when computing the annual deforestation rates in %/yr for these two periods. In this case, we only considered the cover change and computed the ratio $F_{t2}-F_{t1})/F_{t1}$ considering only the pixels that were not covered by clouds in 1973. Computation were done at the national scale and for each forest type in Madagascar: the moist forest in the East, the dry forest in the West, the spiny forest in the South and mangroves on the West coast. To define the forest type, we used a map provided by the MEFT (_"Ministère de l'Environnement et des Forêts à Madagascar"_) showing the boundaries of the four ecoregions in Madagascar. Ecoregions were defined on the basis of climatic and vegetation criteria using the climate classification by @Cornet1974 and the vegetation classification from the 1996 IEFN national forest inventory [@IEFN1996]. Because mangrove forests are highly dynamic ecosystems that can expand or contract on decadal scales depending on environmental factors [@Armitage2015], a fixed delimitation of the mangrove ecoregion might not be appropriate. As a consequence, our estimates of the forest-cover and deforestation rates for mangroves in Madagascar must be considered with caution.

### 2.3 Comparing forest-cover and deforestation rate estimates with previous studies

We compared our estimates of forest-cover and deforestation rate with three previous studies at the national scale for Madagascar [@Harper2007; @MEFT2009; @ONE2015]. @Harper2007 provided forest-cover estimates for the periods _c._ 1953-_c._ 1973-1990-2000, @MEFT2009 provided estimates for the periods 1990-2000-2005 and @ONE2015 provided estimates for the periods 2005-2010-2013. To be able to compare on the same time-periods, we thus considered three additional time-periods to compute forest-covers and deforestation rates from the results our study: 2000-2005, 2005-2010 and 2010-2013. All previous studies provide annual deforestation rate estimates in %/yr. The way annual deforestation rates in %/yr have been computed in these previous studies is not always detailed and might slightly differ from one study to another [see @Puyravaud2003]. But these estimates always correct for the potential presences of residuals clouds on forest maps. @Harper2007 also provide deforestation rate estimates in ha but only for the two periods 1973-1990 and 1990-2000. We converted these estimates into annual deforestation rate estimates in ha/yr. When annual deforestation rate estimates in ha/yr were not reported (for 1953-1973 in @Harper2007 and in @MEFT2009 and @ONE2015), we computed them from the provided forest-cover estimates in each study. These estimates do not correct for the potential presence of residual clouds. Forest-cover and deforestation rates were compared between studies for the whole Madagascar and for the four ecoregions. The same ecoregion boundaries as in our study have been used in the @ONE2015 report but this was not the case for the @Harper2007 article and the @MEFT2009 report, which can explain some of the differences between studies in the forest-cover areas and deforestation rates by ecoregions.

@@

```{r defor, results="asis"}
defor.df <- read.table(file="outputs/defor.txt", sep="\t", header=TRUE)
perc.1 <- round(100*(defor.df$area[defor.df$Year==1973]-defor.df$area[defor.df$Year==2014])/defor.df$area[defor.df$Year==1973])
perc.2 <- round(100*(defor.df$area[defor.df$Year==1953]-defor.df$area[defor.df$Year==2014])/defor.df$area[defor.df$Year==1953])
d.1 <- defor.df$theta[defor.df$Year==1990]
d.2 <- defor.df$theta[defor.df$Year==2010]
tab.1 <- defor.df
names(tab.1) <- c("Year", "Forest (ha)", "Clouds (ha)", "Annual defor. (ha/yr)", "theta (%/yr)")
kable(tab.1)
```

@@

```{r fcc_comp, results="asis"}
fcc.comp <- read.table(file="outputs/fcc_comp.txt", sep=" ", header=TRUE)
tab.2 <- fcc.comp
tab.2 <- tab.2[,c(2,1,3:10)]
tab.2$ForestType <- c("Total",rep("",3),"Moist",rep(" ",3),"Dry",rep(" ",3),"Spiny",rep(" ",3),"Mangroves",rep(" ",3))
col.names <- c("Forest type","Source","1953","1973","1990","2000","2005","2010","2013","2014")
kable(tab.2,col.names=col.names,align=c("l","l",rep("r",8)))
#print(xtable(tab.2,col.names=col.names,align=c("llrrrrrrrrr")),type="html")
```

@@

Table 3: **Comparison of Madagascar annual deforestation rates with previous studies on the period 1953-2014**. Annual deforestation rates are providing in ha/yr and in %/yr (second number in parenthesis). Exact same numbers as in scientific articles and reports from previous studies have been reported for deforestation rate in %/yr. The way annual deforestation rates in %/yr have been computed in these previous studies can slightly differ from one study to another but estimates always correct for the potential presences of clouds on forest maps. On the contrary, deforestation rates in ha/yr have been recomputed from forest-cover estimates in Tab. 2 and do not correct for the potential presence of clouds.

```{r frag, results="asis"}
frag.df <- read.table(file="outputs/frag.txt", sep="\t", header=TRUE)
tab.4 <- frag.df
p.1 <- round(frag.df$interior[frag.df$Year==1973])
p.2 <- round(frag.df$interior[frag.df$Year==2014])
p.3 <- round(frag.df$patch[frag.df$Year==2014]+frag.df$transitional[frag.df$Year==2014])
names(tab.4) <- c("Year","Forest (ha)", "patch (%)", "transitional (%)", "edge (%)", "perforated (%)", "interior (%)", "NA (%)")
kable(tab.4)
```
